---
title: "TSA Final Exam"
author: "Mary Esther Nevener"
date: "May 5, 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## TIME SERIES ANALYSIS FINAL EXAM
###PROBLEM 1
1. [20pts] Consider the MA model $Z_T=0.5+\epsilon_t+\epsilon_{t-1}+\epsilon_{t-2}$
a.  Compute the theoretical ACF and plot it
b.  Generate $Z_t$ for $1\leq t \leq 1000$ and plot $Z_t$
c.  Compute the empirical ACF using the samples and compare it with your answer from (a)

####1a) 
$$
Z_T=0.5+\epsilon_t+\epsilon_{t-1}+\epsilon_{t-2}\\
\mu=0.5, \theta_1=1,  \theta_2 =1\\
\gamma_0=(1+\theta_1^2+\theta_2^2) \sigma^2 = 3 \sigma^2\\
\gamma_1=  (\theta_1+\theta_2\theta_1)\sigma^2 =2\sigma^2\\
\gamma_2=\theta_2\sigma^2 =\sigma_2\\
\rho_1=\frac{\theta_1 + \theta_1\theta_2}{1+\theta_1^2+\theta_2^2}=\frac{2}{3}\\
\rho_2=\frac{\theta_2}{1+\theta_1^2+\theta_2^2}=\frac{1}{3}\\
$$

```{r}
acfma2=ARMAacf(ma=c(1,1), lag.max=10)
acfma2
lags=0:10
plot(lags,acfma2,xlim=c(1,10), ylab="r",type="h", main = "Theoretical ACF for MA(2) with theta1 = 1,theta2=1")
abline(h=0)
``` 


####1b) 
```{r}
zc=arima.sim(n=1000, list(ma=c(1, 1)))
zt=zc+.5
plot(zt, type="b", main = "Simulated MA(2) Series")
``` 
 
####1c) 

##### There are two statistically significant spikes at lags 1 and 2 followed by non-significant values for other lags. However, due to sampling error, the sample ACF did not match the theoretical pattern exactly.

```{r}
acf(zt, xlim=c(1,10), main="ACF for simulated MA(2) Data")
``` 
###PROBLEM 2
2. [20pts] Download 1 time series containing trend and seasonality from http://www.statsci.org/datasets.html
a. Plot each series and compute its ACF and PACF
b. Based on the ACF and PACF, pick 2 distinct models and estimate their parameters
c. Compare the residual for each model
d. Which model is better and why?
###PROBLEM 3
3. [20pts] Derive an expression for the power spectral density,$s_y(\omega)$ for the ARMA(1,1) model nad plot $s_y(\omega)$ for $0 \leq \omega \leq \pi$
###PROBLEM 4
4. [20pts] Prove that conditional expectation minimizes the mean square error.
###PROBLEM 5
5. [20pts] Compute expressions for the partial autocorrelation function for AR(1) and MA(1) models.


 
